 

### 2021-04-17 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

The strength of this buffering varies across forests, and increases with canopy cover, leaf area index ($LAI$) and water availability [@davis_microclimatic_2019]. Diurnally, tropical, temperate broadleaf and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al. 2021; @davis_microclimatic_2019; @zellweger_seasonal_2019]. In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. On the other hand, open-forests with sparse stands, such as needle-leaf forests, have lower buffering capacity, greater vertical air mixing and greater light permeability through canopy layers. For this reason, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999]. During the night, these pine forests have been observed to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a]. Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer. As a result, taller trees up to 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature in the understory. However, beyond a certain tree height, temperature and VPD offset plateaus [@juckerCanopyStructureTopography2018; @hardwickRelationshipLeafArea2015].

@@

Along with temperature, forest canopies moderate relative humidity (RH) in below-canopy microclimates, which tends to decrease with increasing maximum temperature. In three European forests (Broadleaved, Non-Pine Conifer, Pine), canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area.RH moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].

@@

**(paragraph on radiation as a driver)**
Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$. On the other hand, under conditions of constant radiation, increase in $T_{air}$ can induce evaporative cooling [Drake et al. 2018] or under high $T_{air}$, can cause leaf heat stress [Fauset et al. 2018]

@@

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
Leaf thermoregulation is also a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance that facilitates heat dissapation and leaf water content that affects the rate of leaf warming [@fauset_differences_2018; *Koch et al. 1994*]. Leaves with open stomata can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, *with coolest leaves at intermediate sizes (~10mm)*. 
Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance.  However, in circumstances such as during the midday, $T_{leaf}$ can become highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low. This is caused by increase in air temperature and VPD, which limits stomatal conductance along with decreased leaf water content and leaf CO~2~ [*Koch et al. 1994*; @fauset_differences_2018]. On the other, shaded understory leaves under closed canopies should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.

**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** 
*Smaller leaves have higher $g_{a}$ to heat transfer, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017].  
While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature [@fauset_differences_2018]. Large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open [@leighInfluenceLeafSize2017], and significantly hotter under high radiation with stomata closed.

@@

**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`), leaf traits (see following section), and *evapotranspiration* (*where is this discussed? and do we say anything about evaporation?*) together shape in-situ leaf temperatures according to the principles outlined above (Fig. `r fig_leaf_T`).** 
Leaf processes and leaf temperature at a vertical gradient are influenced by vertical microclimate conditions. While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.

Particularly, in open forests with sparse distribution of trees, such as in some needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of increased vertical air mixing, heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. Greater heat stress and $T_{air}$ in the lower canopy is also observed in Australian desert trees [Curtis et al.] and among timberline krummholz needle-leaf trees growing close to the ground 
[@hadleyInfluenceKrummholzMat1987].  Similarly, during the dry season in a semi-decidudous tropical moist forest in Panama where some trees loss canopy leaves, sun leaves can be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016]. Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

@@

In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*] such that gas exchange is strongly constrained by at the tops of(~100 m. tall *Sequoia sempervirens*; *Ambrose et al. 2009, 2010*).
Under constant light conditions (photon flux density), increase in height (>35m) showed decrease in canopy stomatal conductance, along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010]. Adaptations to increasing height and hydraulic contraints show that leaf size [@chinLeafAcclimationLight2017] and lobation [@kusiPlasticLeafMorphology2020] tend to decrease, whereas leaf mass per area (at greater than 20m) tends to increase [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].

@@

Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al. 2019; Tserej & Feeley 2021]. Yet, thermal imagery shows that though leaf area can be a predictor for leaf temperature ranges [$T_{range}$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017]. With increases in effective leaf width, boundary layer thickness and resistance increase along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018]. This lowers evaporation from stomatal pores, and in turn transpirational cooling in large leaves [@fauset_differences_2018]. Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017]. Using leaf energy balance models to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al. 2016 found leaf width to be a dominant factor in determining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$. Leaf width and characteristic dimension are generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref]. Additionally, leaf characteristics such as lobes, leaf elongation and serration, which break the entirety of leaf dimension by enhancing a smaller characteristic dimension, all facilitate more efficient heat dissipation. Leaf lobes are deeper and leaf elongation and serration are more pronounced in sun exposed canopies than in the shaded understory [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].

@@

High light stimulates the need for photoprotection in leaves in order to protect internal photosystem II from chronic photoinhibition through enabling D1 protein replacement. Thus, leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis. Xanthophylls are photoprotective pigments comprised of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. These are found to be proportional to within-canopy light gradients at an inter- and intra-species level. Therefore, $VAZ$ is lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [@niinemetsAcclimationHighIrradiance1998]. In neotropical forests, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pools on an area or chlorophyll basis compared to low light leaves that showed greater mass-based chlorophyll, lutein and neoxanthin carotenoids [Mastubara et al. 2009]

-	antioxidant scavenging function- protect against cellular damage
- Koniger et. al. 1995, and  Matsubara et al. 2009

@@

##### *Leaf lifespan* [Elsa: phenology more broadly? - leaf lifespan, fruiting, flowering links to temperature and thermal gradients?]
*Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves. This is bound to affect metabolism, and may affect other aspects of their performance.*

@@

Across ontogeny, while canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in leaf traits, the difference has been largely attriuted to changes in light conditions [Rosendaal et al. 2006], therefore maintaing similar responses to changes in light with increasing height as highlight within species. Under open-light conditions for canopy and understory,in large gaps and open-grown conditons, 
evidence shows that canopy trees showed significant increase in LMA, photosynthetic capcity per area, and stomatal conductance compared to saplings
[Thomas and Winner, 2002]. However, plasticity to light has been shown difference between species [Houter and Pons 2012]

@@

- Elsa: see also H. Poorter, U. Niinemets, L. Poorter, I. J. Wright, R. Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis. New Phytol. 182, 565–588 (2009). -- LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light.

 "Additionally, shorter understory trees that spend the bulk of their lifecycle in the shade experience different pressures than plant that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well" -Martijn

With increasing height and age, competition for light decreases as availability  of light increases in dense forests. However, depending on growth strategies of species occupying the understory, leaf traits might differ due to species inherent light requirements. Early successional species are considered more shade intolerant than late successional species that are more shade tolerant. Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [@niinemetsControversyTraitsConferring2006]. However, the findings from the above review conflicts with  Kitajima 1994 found for tropical seedlings; shade intolerance species grew faster in the sun AND in the shade, but at the cost of higher mortality in the shade. Lawren and Karoru had contrasting findinds, and so far appears that there is a consensus by now.

@@

Leaf metabolism is strongly shaped by leaf temperature, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
However, as we detail below, there are limited studies comparing thermal responses of sun and shade leaves (Table `r table_leaf_metabolism`).

@@

- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015), however stomatal conductance limitation is also greater, and earlier with canopy height [Sanches et al. 2009] 
Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
-understory leaves open their stomata wider than upper canopy leaves [Kosugi et al. 2021]

@@

**Photosynthesis is generally higher in sun leaves-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020).**
This is primarily driven by the greater light available to sun leaves.

In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma. Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P. (and water availability)
*[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N. Smith [agreed. Again the importance of differentiating between within-species and among species trait variation. Also; steepness of the vertical gradient in light and VPD. In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]*

Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$ or greater in upper canopy.

Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; Slot & Winter 2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure (e.g. Slot & Winter 2017; Smith et al. 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs. e.g. Sage & Kubien 2007; Varhammer et al. 2015 ).

@@

**Leaf thermal tolerance (Tcrit/ T50)**
*(define)*
Typical values ~45-50$^\circ$C, varying somewhat 
across latitude/climate (O’sullivan et al. 2017),
with elevation [@feeley_thermal_2020;  Slot et al unpublished data],
with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012].
Perez and Feeley et al. 2020 found that across species sun leaves with higher photosynthetic heat tolerance experienced higher maximum temperatures and had narrower thermal safety margins. These species may thus be at a greater risk of thermal damage.  
[- Elsa: also see: 
- Perez, T.M. and Feeley, K.J., 2020. Weak phylogenetic and climatic signals in plant heat tolerance. Journal of Biogeography.]

@@

**Similar to photosynthesis, respiration tends to be higher in sun leaves (Chen et al. 2020), but its temperature sensitivity appears to be similar between between sun and shade leaves.**
(Bolstad et al. 1999) did a study at Coweeta, including elev gradient. 
Found higher respiration (at reference T) in canopy leaves (@bolstad_foliar_1999). 
At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999]. (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
Combining reference respiration and $Q_{10}$, @bolstad_foliar_1999 found  a modest net decrease in R for understory leaves relative to canopy leaves as T increases (*see their Fig. 1*)."Higher $Q_{10}$ values in the functional groups of upper canopy where temperatures are the highest may lead to exponential loses of C with increasing global temperatures depending upon the ability of canopy leaves to acclimate" [@cavaleri_foliar_2008]
Thus, @bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.
- Elsa: and see Perez, T.M., Socha, A., Tserej, O. and Feeley, K.J., 2020. PSII heat tolerances characterize thermal generalists and the upper limit of carbon assimilation. Plant, Cell & Environment.

@@

Interaction between degrees of shade x ontogeny x height with rising temperature needs more scientific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient. Leaf traits vary more on the basis on light environment than ontogenetically [@liuPlantSizeEffects2010]

>>>>

The strength of this buffering varies across forests, and increases with canopy cover, leaf area index ($LAI$) and water availability [@davis_microclimatic_2019]. Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al. 2021; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009]. In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. On the other hand, open-forests with sparse stands, have lower buffering capacity, greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [Rambo and NOrth, 2009]. For this reason, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; Rambo and North, 2009]. During the night, pine forests have been observed to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a]. Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer. As a result, taller trees up to 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature in the understory. However, beyond a certain tree height, temperature and VPD offset plateaus [@juckerCanopyStructureTopography2018; @hardwickRelationshipLeafArea2015].

@@

Along with temperature, forest canopies moderate relative humidity (RH) in below-canopy microclimates, which tends to decrease with increasing maximum temperature. In three European forests (Broadleaved, Non-Pine Conifer, Pine), canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area. RH moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].

@@

**(paragraph on radiation as a driver)**
Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$, such that even at low maximum $T_{air}$, leaves exposed to high radiation loads can experience elevated $T_{leaf}$, greater than $T_{air}$. On the other hand, under conditions of constant radiation, increase in $T_{air}$ can induce evaporative cooling [Drake et al. 2018] or under high $T_{air}$, can cause leaf heat stress [Fauset et al. 2018]. Dense canopies that buffer understory tend to absorb and reflect incoming radiation, therefore maintain greater $T_{leaf}$ than $T_{air}$, whereas open-forests, with greater light interception to forest floors, in come cases, reflect a less pronounced gradient of leaf temperatures, and in other cases maintain greater $T_{leaf}$-$T_{air}$ in the lower canopies [**REF**]

@@

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
Leaf thermoregulation is also a tradeoff between water use strategy and optimum leaf temperature, which is influenced by stomtal conductance that facilitates heat dissapation and leaf water content that affects the rate of leaf warming [@fauset_differences_2018; *Koch et al. 1994*]. Leaves with open stomata can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, *with coolest leaves at intermediate sizes (~10mm)*. 
Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance [Drake et al. 2018].  However, in circumstances such as during the midday, $T_{leaf}$ can become highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low. This is caused by increase in air temperature and VPD, which limits stomatal conductance along with decreased leaf water content and leaf CO~2~ [*Koch et al. 1994*; @fauset_differences_2018]. On the other hand, shaded understory leaves under closed canopies should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.

**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** 
*Smaller leaves have higher $g_{a}$ to heat transfer, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017].  
While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature [@fauset_differences_2018]. Large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open, and significantly hotter under high radiation or $T_{air}$ with stomata closed [@leighInfluenceLeafSize2017].

@@

**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`), leaf traits (see following section), and *evapotranspiration* (*where is this discussed? and do we say anything about evaporation?*) together shape in-situ leaf temperatures according to the principles outlined above (Fig. `r fig_leaf_T`).** 
While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.

Particularly, in open forests with sparse distribution of trees, such as in needle-leaf forests, lower canopy $T_{leaf}$ can be substantially warmer than upper canopy $T_{leaf}$, reflecting microclimate conditions of increased vertical air mixing or heating from the ground [@mullerEvidenceEfficientNonevaporative2021], light permeating through canopy layers and low wind velocity or still air in the lower canopy compared to the upper canopy [@mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987]. Greater heat stress and $T_{air}$ in the lower canopy is also observed in Australian desert trees [Curtis et al.] and among timberline krummholz needle-leaf trees growing close to the ground 
[@hadleyInfluenceKrummholzMat1987].  Similarly, during the dry season in a semi-decidudous tropical moist forest in Panama where some trees loss canopy leaves, sun leaves can be cooled relative to $T_{air}$ more than lower canopy shade leaves [@rey-sanchez_spatial_2016]. Under other conditions $T_{leaf}$ tracks $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

@@

In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*] such that gas exchange is strongly constrained by at the tops of(~100 m. tall *Sequoia sempervirens*; *Ambrose et al. 2009, 2010*).
Under constant light conditions (photon flux density), increase in height (>35m) showed decrease in canopy stomatal conductance, along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010]. Adaptations to increasing height and hydraulic contraints show that leaf size [Kaare H. Jensen and Maciej A. Zwieniecki, 2013] and lobation [@kusiPlasticLeafMorphology2020] tend to decrease, whereas leaf mass per area (at greater than 20m) tends to increase [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].

@@

Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al. 2019; Tserej & Feeley 2021]. Yet, thermal imagery shows that effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant than leaf area[@leighInfluenceLeafSize2017]. With increases in effective leaf width, thermal time constant also increases due to increase in boundary layer thickness and resistance, along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018]. Therefore, the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves. Small leaves are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017]. Evidence shows that leaf temperature increased with effective leaf width [fauset et al. 2018]. However, leaf dissection such as lobes, leaf elongation and serration, enhancing a smaller characteristic dimension, thus facilitate efficient heat dissipation, and are evidently characteristics of sun exposed leaves than shaded leaves along a vertical gradient [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006]. Other morphological characteristics such as increased petiole length, characteristic of leaves growing in high-light conditions and in exposed canopy positions, also contribute to heat dissapation [Niinemetes]

@@

High light stimulates the need for photoprotection in leaves in order to protect internal photosystem II from chronic photoinhibition. Thus, leaves exposed to greater irradiance have greater pools of xanthophyll pigments, which are photoprotective pigments comprised of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. These are found to be proportional to within-canopy light gradients at an inter- and intra-species level. Therefore, $VAZ$ is lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [@niinemetsAcclimationHighIrradiance1998]. In neotropical forests, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pools on an area or chlorophyll basis compared to low light leaves that showed greater mass-based chlorophyll, lutein and neoxanthin carotenoids [Mastubara et al. 2009]

-	antioxidant scavenging function- protect against cellular damage

@@

##### *Leaf lifespan* [Elsa: phenology more broadly? - leaf lifespan, fruiting, flowering links to temperature and thermal gradients?]
*Leaf lifespan generally increases with LMA. LMA and leaflifespan is greater with available light, and largely depends on cost of carbon investment in a leaf, which varies across species [**double check**] Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves. This is bound to affect metabolism, and may affect other aspects of their performance.*

@@

Across ontogeny, canopy and understory trees of same species in tropical, temperate, and conifer forests show significant differences in leaf traits, however, the difference has been largely attributed to changes in light conditions rather than ontogeny [Rosendaal et al. 2006], therefore maintaing similar responses to changes in light with increasing height such as in within-species. Under constant light conditions of open-grown canopy and understory, for example in large gaps, evidence shows that canopy trees showed significantly greater LMA, photosynthetic capcity per area, and stomatal conductance compared to saplings, attributing the differences in traits to ontogeny [Thomas and Winner, 2002; Houter and Pons 2012]. On the basis of phenology, temperate understory trees have been observed to leaf out earlier than canopy due to ontogenetic variation in tree size rather than biophysical conditions [Vitasse et al. 2013], however, increase in spring temperatures has shown to have implications on understory phenology, relative to canopy trees [Zohnner and Renne, 2019]

With increasing height and age, competition for light decreases as availability  of light increases in dense forests. However, depending on growth strategies of species occupying the understory, leaf traits might differ due to species inherent light requirements. Early successional species are considered more shade intolerant than late successional species that are more shade tolerant. Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [@niinemetsControversyTraitsConferring2006]. However, the findings from the above review conflicts with  Kitajima 1994 who found for tropical seedlings; shade intolerance species grew faster in the sun and in the shade, but at the cost of higher mortality in the shade. Lawren and Karoru had contrasting findinds, and so far appears that there is no a consensus?

@@

@@

Leaf metabolism is strongly shaped by leaf temperature, and by the traits and environmental drivers reviewed above, all of which vary across vertical forest gradients (Fig. `r fig_schematic`). 
However, as we detail below, there are limited studies comparing thermal responses of sun and shade leaves and vertical gradient (Table `r table_leaf_metabolism`).

@@

- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015), however stomatal conductance limitation is also greater, and earlier with greater canopy height [Sanches et al. 2009] 
Max transpiration rate - increases with canopy height (Kenzo et al. 2015), and quickly decreases with increasing VPD at greater heights [Scaffer].
-understory leaves open their stomata wider than upper canopy leaves [Kosugi et al. 2021]. Stomatal apeture, also controls heat dissapation through evaporative cooling [Taiz and Zeiger]. Stomatal opening is greater in understory compared to canopy

@@

**Photosynthesis is generally higher in sun leaves-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020).**
This is primarily driven by the greater light available to sun leaves. Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$ or greater in upper canopy. Therefore, in dense canopies and across species, photosynthesis is greater at canopy tops where light is most available, compared to middle and lower canopy [REF]. However, light acclimated shade leaves have greater light use efficiency and therefore are observed to maintain higher photosynthetic capacity on a mass-basis?

In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma. Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P. (and water availability)

Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; Slot & Winter 2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure, accompanied by an increase in leaf temperature in dense upper canopies tends to decrease phtotosynthesis [REF](e.g. Slot & Winter 2017; Smith et al. 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs. e.g. Sage & Kubien 2007; Varhammer et al. 2015 ).

*[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N. Smith [agreed. Again the importance of differentiating between within-species and among species trait variation. Also; steepness of the vertical gradient in light and VPD. In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]*

@@

**Leaf thermal tolerance (Tcrit/ T50)**
*(define)*
Typical values ~45-50$^\circ$C, varying somewhat 
across latitude/climate (O' sullivan et al. 2017),
with elevation [@feeley_thermal_2020;  Slot et al unpublished data],
with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012].
Tcrit is greater for buffered leaves that maintain greater effective leaf with [Slot et al. 2021]
Perez and Feeley et al. 2020 found that across species sun leaves with higher photosynthetic heat tolerance experienced higher maximum temperatures and had narrower thermal safety margins. These species may thus be at a greater risk of thermal damage.  
[- Elsa: also see: 
- Perez, T.M. and Feeley, K.J., 2020. Weak phylogenetic and climatic signals in plant heat tolerance. Journal of Biogeography.]

@@

**Similar to photosynthesis, respiration tends to be higher in sun leaves (Chen et al. 2020), but its temperature sensitivity appears to be similar between between sun and shade leaves.**
(Bolstad et al. 1999) did a study at Coweeta, including elev gradient. Conflicting $Q_10$ findings shows no definite trend.
Found higher respiration (at reference T) in canopy leaves (@bolstad_foliar_1999). 
At Coweeta, Q10 showed a variable pattern, with a slight tendency to increase down the canopy [@bolstad_foliar_1999]. (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*)
Combining reference respiration and $Q_{10}$, @bolstad_foliar_1999 found  a modest net decrease in R for understory leaves relative to canopy leaves as T increases (*see their Fig. 1*)."Higher $Q_{10}$ values in the functional groups of upper canopy where temperatures are the highest may lead to exponential loses of C with increasing global temperatures depending upon the ability of canopy leaves to acclimate" [@cavaleri_foliar_2008]
Thus, @bolstad_foliar_1999 found evidence of acclimation, but there remained a declining trend of respiration with elevation.
- Elsa: and see Perez, T.M., Socha, A., Tserej, O. and Feeley, K.J., 2020. PSII heat tolerances characterize thermal generalists and the upper limit of carbon assimilation. Plant, Cell & Environment.

@@

Interaction between degrees of shade x ontogeny with rising temperature needs more scientific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient.
